---
title: "BST 219: Core Principles of Data Science"
subtitle: "Data Wrangling I"
output:
  html_document:
    css: crimson.css
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 3
    number_sections: true
    df_print: paged
    code_folding: show
    highlight: tango
---

```{r setup, include=FALSE}
# This is the "setup" chunk. It runs first and stays hidden (include=FALSE) from
# the final document. We use it to set options that apply to every chunk below.
knitr::opts_chunk$set(
  echo = TRUE,        # Show the R code in the final document
  message = FALSE,    # Hide package start-up messages (keeps output tidy)
  warning = FALSE,    # Hide warnings (keeps output tidy for beginners)
  comment = "#>"      # Prefix printed output with "#>" so it's easy to spot
)
```

# Introduction {#introduction}

Real public health data almost never arrives ready to analyze. Columns are in the wrong
place, the variable you actually need (like a *rate*) hasn't been calculated yet, you want
only a subgroup of rows, or the whole table is shaped in a way that fights the question you're
asking. The work of getting messy data into a clean, analysis-ready form is called **data
wrangling**, and analysts routinely spend the *majority* of a project on it.

This document introduces the two workhorse tools of the **tidyverse** for wrangling:

- **`dplyr`** — for *transforming* data: adding new columns, keeping certain rows, choosing
  columns, sorting, and computing summaries (including summaries within groups).
- **`tidyr`** — for *reshaping* data: converting between "wide" and "long/tidy" layouts, and
  splitting or combining columns.

Along the way you'll meet the **pipe** (`%>%`), which strings small steps into a readable
recipe, and you'll see how to read data files into R.

**Why this matters for public health.** Almost every analysis you'll do — comparing rates
across states, tracking a measure over time, summarizing an outcome by group — starts with
these exact operations. Learning the handful of verbs below gives you a vocabulary for
turning raw data into evidence.

**What you'll learn:**

1. `mutate()`, `filter()`, `select()` — add columns, keep rows, choose columns.
2. `summarize()` and `group_by()` — collapse data into summaries, overall or per group.
3. `arrange()` and `slice()` — sort rows and pick rows by position.
4. `pivot_longer()` and `pivot_wider()` — reshape between wide and tidy layouts.
5. `separate()` and `unite()` — split one column into many, or combine many into one.

<div class="callout tip">
<span class="label">How to use this document</span>
Read the short explanation, run the code chunk below it, then compare the output to what the
explanation predicted. Every line of code has a <code>#</code> comment. Try changing a column
name or a threshold and re-running — experimenting is the fastest way to learn.
</div>

<div class="callout tip">
<span class="label">Packages &amp; styling</span>
This tutorial uses <code>dslabs</code>, <code>dplyr</code>, and the broader
<code>tidyverse</code>. If you haven't installed them, run once in your console:
<code>install.packages(c("dslabs", "tidyverse"))</code>. The crimson look comes from a
companion file, <code>crimson.css</code> — keep it in the same folder as this
<code>.Rmd</code> when you knit.
</div>

# Part 1 — Transforming Data with `dplyr` {#part-1}

The `dplyr` package gives you a small set of **verbs**, each doing one clear job to a data
frame. You'll combine them with the pipe to build up an analysis one readable step at a time.

## The pipe and `mutate()`: adding a column {#mutate}

The `%>%` symbol is the **pipe**. It takes whatever is on its *left* and feeds it as the first
input to the function on its *right*. Read it out loud as the word **"then."** So
`murders %>% mutate(...)` reads as *"take `murders`, **then** add a new column."*

The verb `mutate()` **adds a new column**. Below we create `murder_rate` — murders per
100,000 people — which lets us compare states of very different sizes on a fair footing.

```{r load-and-mutate, message=FALSE}
library(dslabs)     # dslabs provides the practice datasets used throughout
data(murders)       # Load the U.S. gun murders dataset (2010)
head(murders)       # Peek at the first 6 rows: state, abb, region, population, total

library(dplyr)      # dplyr provides the data-transformation "verbs" used below

# mutate() ADDS a column. Here: murder_rate = murders per 100,000 people.
murders <- murders %>% mutate(murder_rate = total / population * 100000)
head(murders)       # The table now has an extra murder_rate column on the right
```

## `filter()`: keeping rows that meet a condition {#filter}

Where `mutate()` works on columns, `filter()` works on **rows**: it keeps only the rows where
a condition is `TRUE`. Below we keep the states whose murder rate is at or below an example
threshold of 0.71.

```{r filter-rows}
# filter() KEEPS only rows where the condition is TRUE.
murders %>% filter(murder_rate <= 0.71)   # Keep states with a low murder rate
```

<div class="callout tip">
<span class="label">Watch out</span>
To <em>test</em> whether two things are equal, use double equals <code>==</code> (e.g.
<code>region == "West"</code>). A single <code>=</code> means "assign," and using it here is a
very common beginner error.
</div>

## `select()`: choosing columns {#select}

`select()` chooses **columns** by name. Pairing `select()` (columns) with `filter()` (rows)
lets you zoom into exactly the slice of data you care about.

```{r select-columns}
# select() chooses columns; filter() chooses rows. Chain them with the pipe.
murders %>%
  select(state, region, murder_rate) %>%   # Keep just these three columns
  filter(murder_rate <= 0.71)              # Then keep only the low-rate states
```

## `summarize()`: collapsing data to a summary {#summarize}

`summarize()` **collapses many rows into a single summary row**. Here we compute one overall
U.S. murder rate. Notice we sum the totals and sum the populations *first*, then divide — this
is the correct national rate.

```{r summarize-total}
# summarize() collapses the whole table into one summary value.
us_murder_rate <- murders %>%
  summarize(murder_rate = sum(total) / sum(population) * 100000)

us_murder_rate   # A 1-row table (a "tibble") holding the single value
```

<div class="callout keyidea">
<span class="label">Key idea (and a common trap)</span>
The national rate is <em>total murders ÷ total population</em>, <strong>not</strong> the
average of the 50 state rates. Averaging the state rates would give small and large states
equal weight and produce a different (wrong) number. This distinction matters constantly in
epidemiology.
</div>

The result above is a small *table*. Sometimes you want the bare *number* instead. Inside a
pipe, the dot `.` means "the data coming from the left," so `.$murder_rate` pulls that one
column out as a plain value.

```{r summarize-pull}
# Pull the single value OUT of the 1x1 table using .$columnname
us_murder_rate <- murders %>%
  summarize(murder_rate = sum(total) / sum(population) * 100000) %>%
  .$murder_rate

us_murder_rate   # Now just a number, not a table
```

<div class="callout tip">
<span class="label">Cleaner alternative</span>
Modern <code>dplyr</code> offers <code>pull(murder_rate)</code>, which does the same thing as
<code>.$murder_rate</code> but reads more clearly. Both are fine; you'll see both in the wild.
</div>

## `group_by()`: summaries within groups {#group-by}

`group_by()` quietly splits the data into groups; a following `summarize()` then runs **once
per group**. Together they answer questions of the form *"what's the value for each ___?"*

```{r group-by}
# group_by() forms groups; summarize() then computes one value PER group.
murders %>%
  group_by(region) %>%                          # One group per region
  summarize(median_rate = median(murder_rate))  # Median murder rate within each region
```

<div class="callout keyidea">
<span class="label">Key idea</span>
<code>group_by()</code> changes nothing you can see on its own — it just tags the rows with
their group. The effect appears only when the next verb (usually <code>summarize()</code>)
runs separately for each group.
</div>

## `arrange()`: sorting rows {#arrange}

`arrange()` **sorts** the rows. By default it sorts smallest-to-largest; wrap a column in
`desc()` to sort largest-to-smallest.

```{r arrange-sort}
# arrange() sorts ascending (smallest first) by default.
murders %>%
  arrange(murder_rate) %>%   # Lowest murder rates first
  head()                     # Show just the first 6 rows

# desc() flips it to descending (largest first).
murders %>%
  arrange(desc(murder_rate)) %>%   # Highest murder rates first
  head()
```

## `slice()`: selecting rows by position {#slice}

`slice()` picks rows by their **position number** (as opposed to `filter()`, which picks rows
by a condition).

```{r slice-rows}
murders %>% slice(1:10)            # Rows 1 through 10
murders %>% slice(-c(3:5, 9:51))   # A minus sign DROPS rows: remove 3–5 and 9–51
murders %>% slice_head(n = 3)      # The first 3 rows
murders %>% slice_tail(n = 5)      # The last 5 rows
```

# Part 2 — Reshaping Data with `tidyr` {#part-2}

Transforming data (Part 1) keeps the table's shape and changes its contents. **Reshaping**
changes the *shape itself* — how rows and columns are laid out. The goal is usually **tidy
data**.

<div class="callout keyidea">
<span class="label">What is "tidy data"?</span>
Data is <strong>tidy</strong> when (1) each <em>variable</em> is its own column, (2) each
<em>observation</em> is its own row, and (3) each <em>value</em> sits in its own cell. Most
tidyverse tools — and most plotting — assume tidy data, so reshaping messy data into this form
is often the first real step of an analysis.
</div>

## The tidyverse and a first look {#tidyverse}

Instead of loading packages one at a time, `library(tidyverse)` loads the whole family at once
(`dplyr`, `tidyr`, `readr`, `ggplot2`, and more). Here we also load the `gapminder` dataset —
health and income indicators by country and year — which we'll use shortly.

```{r load-tidyverse, message=FALSE}
library(tidyverse)   # Loads dplyr, tidyr, readr, ggplot2, and friends in one line
library(dslabs)
data("gapminder")    # Load the gapminder dataset
head(gapminder)      # First 6 rows
```

## Reading data into R {#reading-data}

`read_csv()` (from the `readr` part of the tidyverse) reads a comma-separated file into R. The
file can live on your computer or, as here, on the web — in which case you need an internet
connection for the chunk to run.

```{r read-csv-url, message=FALSE}
library(tidyverse)
library(dslabs)
# The file lives online, so this chunk needs an internet connection to run.
filename <- "https://raw.githubusercontent.com/rafalab/dslabs/master/inst/extdata/fertility-two-countries-example.csv"
wide_data <- read_csv(filename)   # Download and read the CSV into a tibble
head(wide_data)                   # Note: each YEAR is its own column — this is "wide" data
```

## File paths and the working directory {#file-paths}

R reads and writes files relative to a folder called the **working directory**. `getwd()`
shows you where that currently is.

```{r getwd}
# getwd() prints your current WORKING DIRECTORY — where R looks for files by default.
getwd()
```

To read a file that ships *inside* a package, `system.file()` finds that package's folder on
your machine, and `file.path()` joins folder and file name together with the correct slash for
your operating system (so your code works on Windows, Mac, and Linux alike).

```{r file-paths-build}
# system.file() locates a folder inside an installed package on YOUR computer.
path <- system.file("extdata", package = "dslabs")   # dslabs' example-data folder
path
filename <- "murders.csv"                 # Just the file's name
fullpath <- file.path(path, filename)     # Join folder + file into one full path
fullpath
```

## `pivot_longer()`: from wide to tidy {#pivot-longer}

The `wide_data` above has one column per year — convenient for a spreadsheet, but not tidy.
`pivot_longer()` **stacks** those many columns into just two: one holding the old column names
(the years) and one holding the values (fertility). This turns wide data into long/tidy data.

```{r pivot-longer-1}
# pivot_longer() gathers many columns into two (a name column + a value column).
# Here the year columns `1960` through `2015` become rows.
new_tidy_data <- wide_data %>%
  pivot_longer(`1960`:`2015`, names_to = "year", values_to = "fertility")
head(new_tidy_data)
```

<div class="callout tip">
<span class="label">Why the backticks?</span>
Column names that start with a number (like <code>1960</code>) must be wrapped in
back-ticks — <code>`1960`</code> — so R reads them as a name and not as the number 1960.
</div>

The same result can be written by saying which column to *keep* rather than which to *gather*.
A minus sign means "everything except": `-country` gathers every column but `country`.

```{r pivot-longer-2}
# Same result, said differently: gather ALL columns EXCEPT country.
new_tidy_data <- wide_data %>%
  pivot_longer(-country, names_to = "year", values_to = "fertility")
head(new_tidy_data)
```

Because the years came from column *headers*, R stored them as text. We convert `year` to
whole numbers so we can later sort, filter, or plot by year.

```{r pivot-longer-3}
# mutate() + as.integer() converts the "year" text into whole numbers.
new_tidy_data <- wide_data %>%
  pivot_longer(-country, names_to = "year", values_to = "fertility") %>%
  mutate(year = as.integer(year))
head(new_tidy_data)
```

## `pivot_wider()`: from tidy to wide {#pivot-wider}

`pivot_wider()` is the exact **opposite** of `pivot_longer()`: it spreads long data back into
wide form, turning the values of one column into new column headers. This is handy for building
human-readable tables for a report.

```{r pivot-wider}
# pivot_wider() spreads long data back to wide: each year becomes its own column again.
new_wide_data <- new_tidy_data %>%
  pivot_wider(names_from = year, values_from = fertility)

new_wide_data %>% select(country, `1960`:`1967`)   # Show country + the first several years
```

## `separate()`: splitting one column into many {#separate}

Sometimes a single column packs in **two** pieces of information — here, column names like
`1960_fertility` combine a year *and* a measurement. `separate()` splits one column into
several using a separator character (here, the underscore `_`).

```{r separate, message=FALSE, warning=FALSE}
# This file stores year AND measurement together in each column name (e.g. "1960_fertility").
path     <- system.file("extdata", package = "dslabs")
filename <- file.path(path, "life-expectancy-and-fertility-two-countries-example.csv")

raw_data <- read_csv(filename)   # Read the local example file
select(raw_data, 1:5)            # Look at the first 5 columns to see the combined names

data <- raw_data %>% pivot_longer(-country)   # Gather everything except country into name/value
head(data)

# separate() splits ONE column into several, using a separator.
data %>% separate(name, c("year", "variable_name"), "_")   # Split on the underscore

data %>% separate(name, c("year", "variable_name"))        # Underscore is the DEFAULT separator too

# Some names have an extra underscore (e.g. life_expectancy). extra = "merge" keeps the
# leftover pieces together in the LAST column instead of dropping them with a warning.
data %>% separate(name, c("year", "name"), extra = "merge")

# Split, then pivot wider so each measurement becomes its own column.
# (Naming the second piece "name" lets pivot_wider() use its defaults.)
data %>%
  separate(name, c("year", "name"), extra = "merge") %>%
  pivot_wider()
```

## `unite()`: combining columns {#unite}

`unite()` is the opposite of `separate()`: it **glues several columns back into one**. Below is
an alternative route to a tidy table — split the names into three pieces (filling blanks on the
right with `NA`), reunite two of them, spread wider, and tidy up the final column name.

```{r unite}
# Split into THREE named pieces; fill = "right" puts NA where a piece is missing.
var_names <- c("year", "first_variable_name", "second_variable_name")
data %>% separate(name, var_names, fill = "right")

# unite() glues columns back together into a single column.
data %>% separate(name, var_names, fill = "right") %>%
  unite(name, first_variable_name, second_variable_name)   # Recombine the two name pieces

# Then reshape wider so each variable becomes its own column...
data %>% separate(name, var_names, fill = "right") %>%
  unite(name, first_variable_name, second_variable_name) %>%
  pivot_wider()

# ...and finally rename the awkward "fertility_NA" column to just "fertility".
data %>% separate(name, var_names, fill = "right") %>%
  unite(name, first_variable_name, second_variable_name) %>%
  pivot_wider() %>%
  rename(fertility = fertility_NA)
```

# Self-Assessment Questions {#self-assessment}

Try each question **before** revealing the answer. They build from foundational to
challenging and use the same packages and datasets as above (`murders`, `gapminder`). Click
**"Show answer"** to check your work.

## Beginner {#beginner}

**Q1.** Use `mutate()` to add a column called `pop_millions` to `murders`, equal to the
population divided by one million.

<details>
<summary>Show answer</summary>

```r
murders %>% mutate(pop_millions = population / 1000000)
```

</details>

**Q2.** Use `filter()` to keep only the states in the `"West"` region.

<details>
<summary>Show answer</summary>

```r
murders %>% filter(region == "West")   # Note the DOUBLE equals for "is equal to"
```

</details>

**Q3.** Use `select()` to keep only the `state` and `murder_rate` columns.

<details>
<summary>Show answer</summary>

```r
murders %>% select(state, murder_rate)
```

</details>

## Intermediate {#intermediate}

**Q4.** Rewrite this nested command using the pipe: `head(filter(murders, murder_rate < 1))`.

<details>
<summary>Show answer</summary>

```r
murders %>%
  filter(murder_rate < 1) %>%
  head()
```

Read it as: take `murders`, **then** keep low-rate rows, **then** show the first six.

</details>

**Q5.** Find the **median** `murder_rate` for each `region`.

<details>
<summary>Show answer</summary>

```r
murders %>%
  group_by(region) %>%
  summarize(median_rate = median(murder_rate))
```

</details>

**Q6.** Sort the states from **highest** to **lowest** `murder_rate` and show the top six.

<details>
<summary>Show answer</summary>

```r
murders %>%
  arrange(desc(murder_rate)) %>%
  head()
```

`desc()` is what flips the sort from ascending to descending.

</details>

## Advanced {#advanced}

**Q7.** `wide_data` has one column per year. Convert it to tidy (long) form with a `year`
column and a `fertility` column, gathering every column except `country`.

<details>
<summary>Show answer</summary>

```r
wide_data %>%
  pivot_longer(-country, names_to = "year", values_to = "fertility")
```

</details>

**Q8.** In your own words, what does "tidy data" mean, and does `pivot_longer()` or
`pivot_wider()` usually move you *toward* it?

<details>
<summary>Show answer</summary>

Tidy data means each **variable** is a column, each **observation** is a row, and each
**value** is a single cell. When years (or other variable values) are spread across many
column headers, `pivot_longer()` gathers them into rows and moves you toward tidy form.
`pivot_wider()` does the reverse — useful mainly for building readable summary tables.

</details>

**Q9.** Using `gapminder`, compute the **mean `life_expectancy` per continent for the year
2010**. (Hint: `filter()`, then `group_by()`, then `summarize()` — and handle missing values.)

<details>
<summary>Show answer</summary>

```r
gapminder %>%
  filter(year == 2010) %>%                 # Keep only year-2010 rows
  group_by(continent) %>%                  # One group per continent
  summarize(mean_le = mean(life_expectancy, na.rm = TRUE))   # Skip NAs
```

`na.rm = TRUE` tells `mean()` to ignore missing values; without it, any `NA` would make the
whole result `NA`.

</details>

# Frequently Asked Questions (FAQ) {#faq}

**What exactly is the pipe (`%>%`)?**
It takes the result on its left and feeds it as the first argument to the function on its
right. Read it as the word "then." So `murders %>% filter(...) %>% summarize(...)` reads
"take murders, then filter, then summarize." It lets you write steps in the order you'd say
them, instead of nesting functions inside-out.

**I've also seen `|>` — is that the same as `%>%`?**
Almost. `|>` is R's newer built-in ("native") pipe and `%>%` comes from the tidyverse. For
everyday wrangling they behave the same way, so use whichever your course prefers. This
tutorial uses `%>%` throughout.

**When do I use `=` versus `==`?**
Use `==` to *test* whether two things are equal (e.g. `region == "West"` inside `filter()`).
Use `=` to set a function's argument (e.g. `names_to = "year"`). Mixing them up is one of the
most common early errors.

**Why do some column names have backticks, like `` `1960` ``?**
R needs backticks around any name that isn't a plain word — names that start with a number
(`` `1960` ``) or contain spaces. The backticks tell R "this whole thing is one column name."

**What's the difference between `dplyr`, `tidyr`, and the `tidyverse`?**
`dplyr` transforms data (mutate/filter/select/summarize/…); `tidyr` reshapes it
(pivot/separate/unite). The `tidyverse` is a bundle that loads both of those *plus* others
(like `readr` for reading files and `ggplot2` for plots) in a single `library(tidyverse)`
call. You don't need to load `dplyr` separately if you've loaded the tidyverse.

**Do I have to reload the libraries in every chunk?**
No. You only need `library(...)` **once per R session** (per knit). Loading it again does no
harm, which is why you'll sometimes see it repeated, but a single load near the top is enough.

**`mutate()` vs `summarize()` — how do I remember which is which?**
`mutate()` **keeps every row** and adds a column. `summarize()` **collapses** the rows into a
smaller summary (one row, or one row per group after `group_by()`). If you want a value for
every observation, mutate; if you want a total/average/median, summarize.

**Why did `read_csv()` from a web link fail for me?**
Reading a file from a URL needs an internet connection at knit time. If you're offline (or
behind a firewall that blocks the site), download the file first and read it from a local path
instead.

**My `group_by()` line seems to do nothing.**
That's expected — `group_by()` only tags rows with their group. You won't see the effect until
the next verb (usually `summarize()`) runs once per group. Also remember to `ungroup()` if you
later want to operate on the whole table again.

**Where can I learn more?**
The tidyverse documentation at <https://www.tidyverse.org> and the RStudio
[data-wrangling cheatsheets](https://rstudio.github.io/cheatsheets/) are excellent one-page
references to keep beside you while you work.

<div class="callout keyidea">
<span class="label">You've got the toolkit</span>
With <code>mutate</code>, <code>filter</code>, <code>select</code>, <code>summarize</code>,
<code>group_by</code>, <code>arrange</code>, <code>slice</code>, and the reshaping verbs
<code>pivot_longer</code>, <code>pivot_wider</code>, <code>separate</code>, and
<code>unite</code>, you can handle the large majority of everyday data-wrangling tasks in
public health. Keep this file as a reference and reach for it often.
</div>
