---
title: "BST 219: Core Principles of Data Science"
subtitle: "Data Wrangling II"
output:
  html_document:
    css: crimson.css
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    toc_depth: 2
    number_sections: true
    df_print: paged
    code_folding: show
    highlight: tango
---

```{r setup, include=FALSE}
# This "setup" chunk runs first and stays hidden (include=FALSE). It sets options
# that apply to every chunk so the whole document looks consistent.
knitr::opts_chunk$set(
  echo    = TRUE,     # Show the R code in the final document
  message = FALSE,    # Hide package/join messages (keeps output tidy)
  warning = FALSE,    # Hide warnings (keeps output tidy)
  comment = "#>"      # Prefix printed output with "#>"
)
```

# Introduction {#introduction}

Real analyses almost never live in a single table. You'll merge case counts with population
figures, demographics with outcomes, or survey responses with the dates they were collected.
And **time itself** — when a case was reported, how long between exposure and diagnosis — is
central to public health. This week builds two essential wrangling skills: **combining** data
frames, and **working with dates and times**.

Upon successful completion of this week, you should be able to:

1. Combine data frames through joins and binding operations.
2. Parse and manipulate dates and times with the `lubridate` package.

**What you'll learn:** the six `dplyr` **join** types (`left_join`, `right_join`, `inner_join`,
`full_join`, `semi_join`, `anti_join`), how to **bind** rows and columns, the **set operations**
(`intersect`, `union`, `setdiff`, `setequal`), and how to read and pick apart dates with
`lubridate`.

<div class="callout tip">
<span class="label">How to use this document</span>
Read the short explanation, run the code chunk, then compare the output to what the explanation
predicted. Every line of code is commented. The join examples especially reward experimenting —
change which rows are in each small table and watch how the result changes. (Keep
<code>crimson.css</code> in the same folder when you knit.)
</div>

# Setup {#setup}

We'll use `dslabs` datasets throughout. Load the packages and data once, here at the top:

```{r load-packages}
library(tidyverse)   # dplyr (joins, binds, set ops), plus the rest of the tidyverse
library(dslabs)      # Practice datasets
library(lubridate)   # Tools for dates and times

data(murders)                    # U.S. gun murders by state
data(polls_us_election_2016)     # Also loads results_us_election_2016 (state-level results)

head(murders)
head(results_us_election_2016)
```

# Combining Data Frames: Joins {#joins}

A **join** combines two tables by matching rows on a shared column (a "key"). The six join
types differ in *which rows they keep*. We'll start with a real example, then use two tiny
tables so the differences are easy to see.

## Two small tables to experiment with {#join-tables}

The differences between join types are clearest on small tables that *don't* perfectly
overlap. We build one table of six states with their populations, and another of six states
with their electoral votes — the two sets of states only partly match.

```{r join-tables}
# tab1: six states and their populations
tab1 <- slice(murders, 1:6) %>%
        select(state, population)
tab1

# tab2: a different set of six states and their electoral votes
tab2 <- slice(results_us_election_2016, c(1:3, 5, 14, 44)) %>%
        select(state, electoral_votes)
tab2
```

## `left_join()` {#left-join}

`left_join()` keeps **every row of the left (first) table** and adds matching columns from the
right table, filling in `NA` where the right table has no match. Here we join the two small tables, matching on the `state` column that both tables share.

```{r left-join}
left_join(tab1, tab2, by = "state")   # Keep all of tab1; add electoral_votes; NA where no match

tab1 %>% left_join(tab2)              # Same result; dplyr auto-detects the shared "state" column
```

<div class="callout tip">
<span class="label">The "Joining by" message</span>
If you omit <code>by = ...</code>, dplyr figures out the shared column(s) for you and prints a
message telling you which it used (hidden here to keep the output clean). It's good practice to
state <code>by</code> explicitly so your intent is clear.
</div>

## `right_join()` {#right-join}

`right_join()` is the mirror image: it keeps **every row of the right table** instead.

```{r right-join}
tab1 %>% right_join(tab2)   # Keep all of tab2; NA where tab1 has no match
```

## `inner_join()` {#inner-join}

`inner_join()` keeps **only rows that appear in both** tables — the intersection of the keys.

```{r inner-join}
inner_join(tab1, tab2)   # Keep only states found in BOTH tables
```

## `full_join()` {#full-join}

`full_join()` keeps **every row from both** tables, filling `NA` wherever a side is missing.

```{r full-join}
full_join(tab1, tab2)   # Keep all states from EITHER table
```

## `semi_join()` {#semi-join}

`semi_join()` is a **filtering join**: it keeps the rows of the left table that *have* a match
in the right table, but it does **not** add any of the right table's columns.

```{r semi-join}
semi_join(tab1, tab2)   # Keep tab1 rows that have a match in tab2 (columns unchanged)
```

## `anti_join()` {#anti-join}

`anti_join()` is the other filtering join: it keeps the left-table rows that have **no** match
in the right table — useful for finding what's *missing*.

```{r anti-join}
anti_join(tab1, tab2)   # Keep tab1 rows that have NO match in tab2
```

<div class="callout keyidea">
<span class="label">Key idea</span>
<strong>Mutating joins</strong> (<code>left</code>, <code>right</code>, <code>inner</code>,
<code>full</code>) combine columns from both tables. <strong>Filtering joins</strong>
(<code>semi</code>, <code>anti</code>) use the second table only to decide which rows of the
first table to <em>keep</em> — they never add its columns.
</div>

# Binding Data Frames {#binding}

Joins match rows by a key. **Binding** is simpler: it just glues tables together by position,
assuming they already line up.

## `bind_cols()`: adding columns {#bind-cols}

`bind_cols()` places tables **side by side**, matching them row-by-row by position.

```{r bind-cols}
# Here tab1 and tab2 are two halves of the SAME rows, split by column.
tab1 <- murders[, 1:2]              # The first two columns
tab2 <- murders[, 3:4]              # The next two columns
new_tab <- bind_cols(tab1, tab2)    # Glue them back together, column-wise
head(new_tab)
```

<div class="callout tip">
<span class="label">Prefer a join when there's a key</span>
<code>bind_cols()</code> trusts that the rows are already in the same order — if they aren't,
you'll silently pair up the wrong records. When the tables share an identifier, a join is
safer because it matches on the key rather than on position.
</div>

## `bind_rows()`: stacking rows {#bind-rows}

`bind_rows()` stacks tables **on top of each other**, matching them up by column name.

```{r bind-rows}
# Now tab1 and tab2 are two different sets of ROWS with the same columns.
tab1 <- murders[1:2, ]      # The first two rows
tab2 <- murders[3:4, ]      # The next two rows
bind_rows(tab1, tab2)       # Stack them into one table
```

# Set Operations {#set-operations}

Set operations treat each table's **rows** (or each vector's **elements**) as a set, and
compare two sets. Each example below shows the idea on plain vectors first, then on data
frames. We'll reuse two overlapping slices of `murders`:

```{r setops-tables}
# Two slices that overlap in rows 3, 4, and 5:
tab1 <- murders[1:5, ]   # Rows 1–5
tab2 <- murders[3:7, ]   # Rows 3–7
```

## `intersect()` {#intersect}

`intersect()` returns the elements (or rows) present in **both** sets.

```{r intersect}
intersect(1:10, 6:15)                       # Numbers in both -> 6, 7, 8, 9, 10
intersect(c("a", "b", "c"), c("b", "c", "d")) # Shared letters -> "b", "c"

intersect(tab1, tab2)                       # Rows present in BOTH tab1 and tab2 (rows 3–5)
```

## `union()` {#union}

`union()` returns all **distinct** elements (or rows) found in **either** set.

```{r union}
union(1:10, 6:15)                       # All numbers, no duplicates -> 1..15
union(c("a", "b", "c"), c("b", "c", "d")) # -> "a", "b", "c", "d"

union(tab1, tab2)                       # All distinct rows across both tables
```

## `setdiff()` {#setdiff}

`setdiff()` returns what's in the **first** set but **not** the second. Order matters here.

```{r setdiff}
setdiff(1:10, 6:15)   # In 1:10 but not 6:15 -> 1, 2, 3, 4, 5
setdiff(6:15, 1:10)   # Reversed -> 11, 12, 13, 14, 15

setdiff(tab1, tab2)   # Rows in tab1 but not tab2 (rows 1–2)
```

## `setequal()` {#setequal}

`setequal()` asks whether two sets contain **exactly the same** elements, ignoring order.

```{r setequal}
setequal(1:5, 1:6)     # FALSE — different elements
setequal(1:5, 5:1)     # TRUE  — same elements, order doesn't matter
setequal(tab1, tab2)   # FALSE — different rows
```

# Dates and Times with `lubridate` {#dates}

Dates are deceptively tricky: they come in dozens of text formats and need real calendar logic
to manipulate. The `lubridate` package makes reading and dissecting them straightforward.

## Dates as a data type {#date-type}

R has a dedicated `Date` type. Underneath, a date is stored as a **number** — the count of days
since January 1, 1970 — which is what lets R do arithmetic with dates.

```{r date-type}
head(polls_us_election_2016$startdate)    # These look like dates...
class(polls_us_election_2016$startdate)   # ...and R confirms they are "Date" objects

as.numeric(polls_us_election_2016$startdate) %>% head()   # Under the hood: days since 1970-01-01
```

## Extracting components of a date {#date-components}

First we grab 10 random dates to work with. `set.seed()` makes the random draw reproducible, so
you get the same 10 dates every time you knit.

```{r sample-dates}
set.seed(2)   # Fix the random seed for reproducibility
dates <- sample(polls_us_election_2016$startdate, 10) %>% sort()   # 10 random dates, sorted
dates
```

The functions `month()`, `day()`, and `year()` pull out each piece of a date.

```{r date-parts}
# Build a small table with each date broken into its parts.
data.frame(month = month(dates),
           day   = day(dates),
           year  = year(dates))
```

Add `label = TRUE` to get the month's **name** instead of its number.

```{r month-label}
month(dates, label = TRUE)   # e.g. "Nov" instead of 11
```

## Parsing dates from messy text {#parsing}

The real power of `lubridate` is turning **text** into proper dates. The `ymd()` function reads
**y**ear-**m**onth-**d**ay order and is remarkably forgiving about inconsistent formatting.

```{r ymd-parse}
# All of these get parsed to the same tidy YYYY-MM-DD dates, despite the mess.
x <- c(20090101, "2009-01-02", "2009 01 03", "2009-1-4",
       "2009-1, 5", "Created on 2009 1 6", "200901 !!! 07")
ymd(x)
```

The same digits mean different dates depending on the order you assume. `lubridate` gives you a
function for each order — pick the one that matches your data. **All dates MUST be in the same order for any of these functions to work properly**.

```{r parse-variants}
x <- "09/01/02"
ymd(x)   # year-month-day
ydm(x)   # year-day-month
myd(x)   # month-year-day
mdy(x)   # month-day-year
dmy(x)   # day-month-year
dym(x)   # day-year-month
```

## Working with times {#times}

`lubridate` handles times and full time stamps too.

```{r times}
now()              # Current date-time in your local time zone
now("GMT")         # ...or in a specified time zone

now() %>% hour()   # Pull out just the hour
now() %>% minute() # ...the minute
now() %>% second() # ...the second

x <- "12:34:56"
hms(x)             # Parse an hours:minutes:seconds string

x <- "Nov/2/2012 12:34:56"
mdy_hms(x)         # Parse a combined date-and-time string (month-day-year, then h:m:s)
```

# Self-Assessment Questions {#self-assessment}

Try each question **before** revealing the answer. They build from foundational to challenging
and use the packages and data loaded above. Click **"Show answer"** to check your work.

## Beginner {#beginner}

**Q1.** Which join keeps **every row of the left table**, filling in `NA` where the right table
has no match?

<details>
<summary>Show answer</summary>

`left_join()`. (`right_join()` does the same but keeps every row of the *right* table.)

</details>

**Q2.** Use a `lubridate` function to turn the text `"2021-03-15"` into a proper date.

<details>
<summary>Show answer</summary>

```r
ymd("2021-03-15")   # year-month-day order
```

</details>

**Q3.** Which function stacks two data frames on top of each other (adding rows), matching by
column name?

<details>
<summary>Show answer</summary>

`bind_rows()`. (`bind_cols()` glues tables side-by-side instead, adding columns.)

</details>

## Intermediate {#intermediate}

**Q4.** In one sentence each, explain the difference between `inner_join()` and `full_join()`.

<details>
<summary>Show answer</summary>

`inner_join()` keeps only the rows whose key appears in **both** tables. `full_join()` keeps
**every** row from **both** tables, filling `NA` wherever one side has no match.

</details>

**Q5.** Given a vector of dates `d`, extract the month as a **name** (like "Mar") rather than a
number.

<details>
<summary>Show answer</summary>

```r
month(d, label = TRUE)
```

</details>

**Q6.** You have `tab1` and `tab2`. Which single join returns the rows of `tab1` that have **no**
match in `tab2`?

<details>
<summary>Show answer</summary>

```r
anti_join(tab1, tab2)
```

An `anti_join` is a filtering join — it keeps rows of the first table with no match in the
second, and adds no new columns.

</details>

## Advanced {#advanced}

**Q7.** Explain the difference between a **mutating join** and a **filtering join**, and give an
example of each.

<details>
<summary>Show answer</summary>

A **mutating join** (`left_join`, `right_join`, `inner_join`, `full_join`) combines the two
tables' **columns**, adding the right table's variables to the left. A **filtering join**
(`semi_join`, `anti_join`) uses the second table only to decide which **rows** of the first
table to keep — it never adds columns. Example: `left_join` adds election columns to the murders
table; `semi_join` just keeps the murders rows that have a matching election record.

</details>

**Q8.** You have a vector of date strings written as month/day/year, e.g., `"11/02/2012"`. Which
`lubridate` function parses them correctly?

<details>
<summary>Show answer</summary>

```r
mdy("11/02/2012")   # month-day-year order
```

Choosing `ymd()` or `dmy()` here would misread the pieces — always match the function to the
order your data is actually in.

</details>

**Q9.** Using a set operation, return only the rows that two data frames `tab1` and `tab2` have
in common.

<details>
<summary>Show answer</summary>

```r
intersect(tab1, tab2)
```

`intersect()` returns the rows present in both tables; `union()` would return all distinct rows,
and `setdiff()` those in one but not the other.

</details>

# Frequently Asked Questions (FAQ) {#faq}

**How do I remember the difference between `left`, `right`, `inner`, and `full` joins?**
They differ only in which rows they keep. `left` keeps all rows of the first table; `right`
keeps all of the second; `inner` keeps only rows matched in both; `full` keeps everything from
both. In all four, the matched columns from both tables are combined.

**What are `semi_join` and `anti_join` for?**
They *filter* rather than combine. `semi_join` keeps the first table's rows that **have** a match
in the second; `anti_join` keeps the rows that **don't**. Neither adds the second table's
columns — they're for questions like "which of my records are missing from this other list?"

**What if the key columns have different names in the two tables?**
Tell the join how they line up with a named vector, e.g. `by = c("state" = "state_name")`. The
left name is the column in the first table, the right name the column in the second.

**Why did my join print a "Joining with `by = ...`" message?**
When you don't specify `by`, dplyr auto-detects the columns the two tables share and reports
which it used. It's harmless, but stating `by` explicitly makes your intent clear and avoids
surprises if the tables share more than one column.

**When should I use `bind_cols` instead of a join?**
Rarely. `bind_cols` pairs rows purely by position and trusts they're already aligned; if they
aren't, you'll silently mismatch records. Whenever the tables share an identifier, use a join so
rows are matched by key, not by luck of ordering.

**Why is my date column being treated as text (or a number)?**
Dates read from a file often come in as character strings. Parse them into real dates with the
`lubridate` function whose letters match your format — `ymd()`, `mdy()`, `dmy()`, and so on —
after which sorting, plotting, and date arithmetic all work correctly.

**How do I choose between `ymd()`, `mdy()`, and `dmy()`?**
Match the function name to the order the pieces appear in your data: `"2021-03-15"` is
year-month-day → `ymd()`; `"03/15/2021"` is month-day-year → `mdy()`; `"15/03/2021"` is
day-month-year → `dmy()`.

**Why do I `set.seed()` before `sample()`?**
`sample()` draws at random, so it gives different results each run. `set.seed()` fixes the
random number generator so the "random" draw is the same every time you knit — essential for a
reproducible document.

<div class="callout keyidea">
<span class="label">You've leveled up</span>
You can now bring separate tables together the right way — matching on a key with joins, or
stacking with binds — compare them with set operations, and turn messy date text into dates you
can compute with. These are the everyday moves of real data work; keep this file handy as a
reference.
</div>
